---
title: "Plotting a map of the Divvy stations"
author: "Peter Carbonetto"
output: html_document
---

<!-- Define defaults shared by all workflowr files. -->
```{r read-chunk, include=FALSE, cache=FALSE}
knitr::read_chunk("chunks.R")
```

<!-- Update knitr chunk options -->
```{r knitr-opts-chunk, include=FALSE}
```

<!-- Insert the date the file was last updated -->
```{r last-updated, echo=FALSE, results='asis'}
```

<!-- Insert the code version (Git commit SHA1) if Git repository
 exists and R package git2r is installed -->
```{r code-version, echo=FALSE, results='asis'}
```

In the previous two episodes, we've done lots of work importing and
manipulating the Divvy data. Now let's do something fun with these
data. we will use the Divvy trip and station data to generate a map of
Chicago.
<br>
<br>

## Our first plot of the Divvy stations

[ggplot2](http://ggplot2.tidyverse.org) has emerged as the most
popular R package for creating visualizations from data.  Although
most people find the interface initially unintuitive, once you get
used to it, it is a very powerful tool, and can create sophisticated
plots in only a few lines of code.

```{r, eval=FALSE}
install.packages("ggplot2")
```

For example, in only one line of code we can create a plot of the
Divvy stations by geographic location (latitude & longitude).

```{r, eval=FALSE}
library(ggplot2)
ggplot(stations,aes(x = longitude,y = latitude)) + geom_point()
```

I'd like to suggest a couple adjustments to the plotting parameters:

```{r, eval=FALSE}
ggplot(stations,aes(x = longitude,y = latitude)) +
  geom_point(shape = 21,fill = "darkblue",color = "white",size = 2.5)
```

ggplot implements transformation and adjustments with separate
functions by modifying the initial plot using the "+" operator.
Here's an example:

```{r, eval=FALSE}
ggplot(stations,aes(x = longitude,y = latitude)) +
  geom_point(shape = 21,fill = "darkblue",color = "white",size = 2.5) +
  theme_minimal() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())
```

## Scale the area of each station by total number of departures

So far, our plot traces the outlines of the City of Chicago and the
Lake Michigan shore. I would like to add an additional piece of
information to this visualization: the number of trips from each
station.

To do this, we need to add this information to the "stations" data
frame. Let's create a new variable containing the counts of the number
of departures at each station:

```{r, eval=FALSE}
counts <- table(trips$from_station_name)
```

Let's double-check that the stations in the "counts" variable are
listed in the same order as the stations in the data frame:

```{r, eval=FALSE}
all(names(counts) == rownames(stations))
```

Looks good, so we can use this variable to create a new column in the
stations data frame:

```{r, eval=FALSE}
counts   <- as.vector(counts)
counts   <- data.frame(departures = counts)
stations <- cbind(stations,counts)
head(stations)
summary(stations$departures)
```

Now that we have the additional "departures" column in the data frame,
we can incorporate these data into the new plot:

```{r, eval=FALSE}
ggplot(stations,aes(x = longitude,y = latitude, size = departures)) +
  geom_point(shape = 21,fill = "darkblue",color = "white") +
  theme_minimal() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())
```

This looks nice, but isn't doesn't accomplish *exactly* what I wanted
because I wanted to scale the *area* of each circle by the number of
departures, not the diameter of each circle. This requires one change
to our plotting code.

```{r, eval=FALSE}
ggplot(stations,aes(x = longitude,y = latitude, size = sqrt(departures))) +
  geom_point(shape = 21,fill = "darkblue",color = "white") +
  theme_minimal() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())
```

Notice that the location of the downtown is now apparent, and that
there generally more departures per station on the North Side
(although Hyde Park has a decent amount of hiking activity).

Do you observe any other interesting trends?

## Visualizing seasonal biking trends

Let's now use the Divvy trip data to examine biking trends in Chicago
over the course of one year.

Convert the "start week" column to a factor, then we can easily count
the total number of Divvy trips per week of the year.

```{r, eval=FALSE}
trips       <- transform(trips,start.week = factor(start.week))
counts.week <- table(trips$start.week)
counts.week <- as.vector(counts.week)
counts.week <- data.frame(week = 0:52,departures = counts.week)
head(counts.week)
tail(counts.week)
```

```{r, eval=FALSE}
ggplot(counts.week,aes(x = week,y = departures)) +
  geom_point(color = "darkblue") +
  geom_line(color = "darkblue") +
  theme_minimal() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())
```

What expected or surprising trends do you notice from this plot? What
improvements would you make to this figure to better highlight the
interesting trends?

## Save your session

Before moving to the next episode, let's save our current session.

```{r, eval=FALSE}
save.image(file = "workshop.RData")
```
<br>

## Recap

In this episode, we illustrated how ggplot can be used. We only
explored the most basic features of ggplot, and it has many, many
more. The key point is that ggplot works very well provided you take
the time beforehand to organize your data into a data frame, and
(e.g., factors).
