---
title: "Create plots from the Divvy station & trip data"
author: "Peter Carbonetto"
output: html_document
---

<!-- Define defaults shared by all workflowr files. -->
```{r read-chunk, include=FALSE, cache=FALSE}
knitr::read_chunk("chunks.R")
```

<!-- Update knitr chunk options -->
```{r knitr-opts-chunk, include=FALSE}
```

<!-- Insert the date the file was last updated -->
```{r last-updated, echo=FALSE, results='asis'}
```

<!-- Insert the code version (Git commit SHA1) if Git repository
 exists and R package git2r is installed -->
```{r code-version, echo=FALSE, results='asis'}
```

Up to this point, we have done lots of work importing and manipulating
the Divvy data. Now let's do something fun with these data. In this
episode, we will use the Divvy trip and station data to generate some
interesting plots.
<br>
<br>

## Our first plot: a map of the Divvy stations

[ggplot2](http://ggplot2.tidyverse.org) has emerged as the most
popular R package for generating data visualizations. Although the
interface can be initially unintuitive, it is a very powerful tool,
and can create surprisingly sophisticated plots in only a few lines of
code.

```{r, eval=FALSE}
install.packages("ggplot2")
library(ggplot2)
```

For example, in only a few lines of code, we can create a plot of the
Divvy stations by geographic location (latitude & longitude).

```{r, eval=FALSE}
ggplot(data    = stations,
       mapping = aes(x = longitude,y = latitude)) +
  geom_point()
```

What geographic features does this plot evoke?

I'd like to suggest a couple adjustments to the plotting parameters. 

```{r, eval=FALSE}
ggplot(data    = stations,
       mapping = aes(x = longitude,y = latitude)) +
  geom_point(shape = 21,fill = "darkblue",color = "white",size = 2.5)
```

You can consult `help(geom_point)` for more information about the
plotting parameters. I also find
[this website](http://ggplot2.tidyverse.org) to be a great reference
on ggplot.

There are many functions in the ggplot2 package for adjusting the plot
setting. All changes to a plot are implemented using the "+" operator.
Here's an example:

```{r, eval=FALSE}
ggplot(data    = stations,
       mapping = aes(x = longitude,y = latitude)) +
  geom_point(shape = 21,fill = "darkblue",color = "white",size = 2.5) +
  theme_minimal() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())
```
<br>

## Scale the area of each station by total number of departures

So far, the plot we have created traces the outlines of the City of
Chicago and Lake Michigan. I would like to add an additional piece of
information to this visualization: the amount of activity at each
station, which should be a rough proxy for population density in
Chicago.

To do this, we need to add this information to the "stations" data
frame because this is the data frame we have been providing as an
input to "ggplot". Let's create a new variable containing the counts of
the number of departures at each station:

```{r, eval=FALSE}
counts <- table(trips$from_station_name)
```

Let's double-check that the stations in the "counts" variable are
listed in the same order as the stations in the data frame:

```{r, eval=FALSE}
all(names(counts) == rownames(stations))
```

This should output TRUE---remember that we were careful to create the
factors in the trips data frame so that the levels were ordered in the
same way as the stations in the stations table. 

Let's use this new variable to create a new column in the stations
data frame:

```{r, eval=FALSE}
counts   <- as.vector(counts)
counts   <- data.frame(departures = counts)
stations <- cbind(stations,counts)
head(stations)
summary(stations$departures)
```

Now that we have the additional "departures" column in the data frame,
we can use these data in our new plot:

```{r, eval=FALSE}
ggplot(data    = stations,
       mapping = aes(x = longitude,y = latitude,
                     size = departures)) +
  geom_point(shape = 21,fill = "darkblue",color = "white") +
  theme_minimal() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())
```

This looks nice, but isn't doesn't accomplish *exactly* what I wanted
because I wanted to scale the *area* of each circle by the number of
departures, and not the diameter of each circle. Because ggplot can
also plot *expressions*, this requires only one change to our plotting
code.

```{r, eval=FALSE}
ggplot(data    = stations,
       mapping = aes(x = longitude,y = latitude,size = sqrt(departures))) +
  geom_point(shape = 21,fill = "darkblue",color = "white") +
  theme_minimal() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())
```

Notice that the location of the downtown is apparent by the higher
density of biking activity. Also observe that there are generally more
departures per station on the North Side of Chicago, and,
interestingly, Hyde Park also appears to have a good amount of biking
activity.

What else interesting do you observe from this plot?
<br>
<br>

## Visualizing seasonal biking trends

Let's now use the Divvy trip data to examine biking trends over the
course of one year in Chicago.

Convert the "start week" column to a factor, then we can easily count
the total number of Divvy trips per week of the year.

```{r, eval=FALSE}
trips       <- transform(trips,start.week = factor(start.week))
counts.week <- table(trips$start.week)
counts.week <- as.vector(counts.week)
counts.week <- data.frame(week = 0:52,departures = counts.week)
head(counts.week)
tail(counts.week)
```

```{r, eval=FALSE}
ggplot(counts.week,aes(x = week,y = departures)) +
  geom_point(color = "darkblue") +
  geom_line(color = "darkblue") +
  theme_minimal() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())
```

What expected or surprising trends do you notice from this plot? What
improvements would you make to this figure to better highlight the
interesting trends?

## Seasonal trends at the University of Chicago

```{r, eval=FALSE}
trips.uofc <- subset(trips,from_station_name == "University Ave & 57th St")
counts.week.uofc <-
  data.frame(week       = 0:52,
             departures = as.vector(table(trips.uofc$start.week)))
```

```{r, eval=FALSE}
ggplot(counts.week.uofc,aes(x = week,y = departures)) +
  geom_point(color = "darkblue") +
  geom_line(color = "darkblue") +
  theme_minimal() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())
```

When we analyze trips taken at the University of Chicago bike station,
the "bump" during warmer months flattens out. This is probably because
a large fraction of University of Chicago students leave during the
summer.

## Save your session

It is time to the current state of our session before moving to the
[next episode](uofc.html).

```{r, eval=FALSE}
save.image(file = "workshop.RData")
```
<br>

## Recap

In this episode, we illustrated how ggplot can be used. We only
explored the most basic features of ggplot, and it has many, many
more. The key point is that ggplot works very well provided you take
the time beforehand to organize your data into a data frame, and
(e.g., factors, dates & times).

In the [final episode](), we will take a closer look at the biking
trends at the University of Chicago.
