---
title: "Reading the Divvy station data into R"
author: "Peter Carbonetto"
output: html_document
---

<!-- Define defaults shared by all workflowr files. -->
```{r read-chunk, include=FALSE, cache=FALSE}
knitr::read_chunk("chunks.R")
```

<!-- Update knitr chunk options -->
```{r knitr-opts-chunk, include=FALSE}
```

<!-- Insert the date the file was last updated -->
```{r last-updated, echo=FALSE, results='asis'}
```

<!-- Insert the code version (Git commit SHA1) if Git repository
 exists and R package git2r is installed -->
```{r code-version, echo=FALSE, results='asis'}
```

## About this file

This file is either a webpage generated from an R Markdown document,
or the original R Markdown file itself.

[R Markdown](http://r4ds.had.co.nz/r-markdown.html) combines
[Markdown](http://daringfireball.net/projects/markdown), a popular,
easy-to-read plain text formatting syntax, with "chunks" of R code. R
Markdown notebooks can be easily rendered into various formats (e.g.,
PDF, HTML) for creating
[narratives](https://www.oreilly.com/ideas/the-state-of-jupyter) of
your code and results.

There are many benefits to documenting your data analyses in R
Markdown---we will not have time to explore all these benefits here.

## How to use this file

We will use this R Markdown document *interactively.*

Although you can easily generate all the output and figures with a
single command from the rmarkdown package, in this workshop we will
walk through the code and text by typing the lines of code in the R
console, or by copying & pasting the code.

## About the Divvy station data

## Loading the Divvy station data

The R environment is where all *variables* and *functions* are stored
and accessed.

Immediately after launching R, your R environment should be empty.

```{r}
ls()
```

We will create a new variable called `station.data` by running
function `read.table`, and assigning the result of `read.table` to
`station.data`:

```{r}
station.data <- read.table
```

What kind of variable is

## What is a data frame?
