---
title: "Setup for analysis of Divvy data"
author: "Peter Carbonetto"
output: html_document
---

<!-- The file analysis/chunks.R contains chunks that define default settings
shared across the workflowr files. -->
```{r read-chunk, include=FALSE, cache=FALSE}
knitr::read_chunk("chunks.R")
```

<!-- Update knitr chunk options -->
```{r knitr-opts-chunk, include=FALSE}
```

<!-- Insert the date the file was last updated -->
```{r last-updated, echo=FALSE, results="asis"}
```

<!-- Insert the code version (Git commit SHA1) if Git repository exists and R
 package git2r is installed -->
```{r code-version, echo=FALSE, results="asis"}
```
<br>

## Setup instructions

```{r, echo=FALSE, results="hide", eval=TRUE, message=FALSE}
library(data.table)
library(lubridate)
library(ggplot2)
```

To work through the lessons on your computer, please follow these
setup instructions first.

1. You have two options for running R or Rstudio Desktop for this
workshop: (a) you can use your computer, or; (b) you can use the RCC
compute cluster. If you use the RCC cluster, I recommend connecting
with
[ThinLinc](https://rcc.uchicago.edu/docs/connecting/#connecting-with-thinlinc).
If you do not have an RCC user account, the instructor can provide you
with a [Yubikey](yubikeys.html).

2. If you are not using the RCC cluster, install R from the
[CRAN website](http://cran.r-project.org) and, optionally,
[RStudio Desktop](https://www.rstudio.com/products/rstudio/download).
*Note:* If you do not use RStudio Desktop, you will need to have
additional software installed for X Windows graphics (e.g.,
[XQuartz](http://xquartz.org) on Mac). Without this, you will not be
able to create plots.

3. Download or clone the
[git repository](https://github.com/rcc-uchicago/R-intro-divvy). Click
[here](https://github.com/rcc-uchicago/R-intro-divvy/archive/master.zip)
to download a Zip file containing all the files. If you are on the
cluster, you can use "wget" or "git clone" to download the git
repository.

4. Download the Divvy data files into the "data" directory, and
prepare the data files for analysis. **Easiest approach:** download
the prepared data files [here](https://uchicago.box.com/v/divvy), and
move the files into the "data" directory. The file on Box is about 82
MB, so it may take a few minutes to download. **More complicated
approach:** download the data directly from the Divvy website. I have
provided a bash script `code/retrieve_divvy_data.sh` to do this
automatically. This script will also move the files to the expected
location, and prepare the files for the analysis in R. This script
reproduces the data processing steps starting with the
[original source](https://www.divvybikes.com/system-data).

5. After completing Step 4, check that you have all the necessary data
files in the "data" directory. For example, on my MacBook Air, I have:

```{r}
list.files("../data")
```

6. Launch R or RStudio Desktop. If you are on the compute cluster, you
need to reserve a compute node, then load an R and/or RStudio
module. For example,

```
sinteractive --time=02:00:00 --mem=2G --partition=broadwl
module load R/3.3
module load rstudio
rstudio
```

7. Make sure your working directory in R or RStudio is set to the
"analysis" directory of the git repository. For example, on my MacBook
Air, the current working directory in R is

```{r}
getwd()
```

If you are not in the analysis directory, use the "setwd" function or
the RStudio menus to change the working directory.

Once you have completed these steps, you are ready to move on to the
lessons.
<br>
<br>

## Session information

The "sessionInfo" function shows the version of R I am using, and the
packages that I used to generate the webpages from the R Markdown
documents.

```{r session-info}
```
