---
title: "Import and inspect the Divvy station data"
author: "Peter Carbonetto"
output: html_document
---

<!-- Define defaults shared by all workflowr files. -->
```{r read-chunk, include=FALSE, cache=FALSE}
knitr::read_chunk("chunks.R")
```

<!-- Update knitr chunk options -->
```{r knitr-opts-chunk, include=FALSE}
```

<!-- Insert the date the file was last updated -->
```{r last-updated, echo=FALSE, results="asis"}
```

<!-- Insert the code version (Git commit SHA1) if Git repository
 exists and R package git2r is installed -->
```{r code-version, echo=FALSE, results="asis"}
```

## About this file

This file is either a webpage generated from an R Markdown document,
or the original R Markdown file itself.

[R Markdown](http://r4ds.had.co.nz/r-markdown.html) combines
[Markdown](http://daringfireball.net/projects/markdown) with "chunks"
of R code. R Markdown notebooks can be easily rendered into various
formats (e.g., PDF, HTML) for creating sharable and reproducible
[narratives](https://www.oreilly.com/ideas/the-state-of-jupyter) of
your data analyses.

There are many benefits to documenting your data analyses in R
Markdown---we will not have time to explore all the benefits here.

## How to use this file

We will use this R Markdown document *interactively.*

Although you can easily generate all the output and figures with a
single command from the
[rmarkdown package](https://cran.r-project.org/package=rmarkdown), in
this workshop we will walk through the data analyses by typing or
copying & pasting the lines of code into the R console.

## About the Divvy station data

After you launch R or RStudio, double-check that your working
directory is set to the "analysis" directory. For example, on my
MacBook Air the current working directory is

```{r, eval=TRUE}
getwd()
```

If it is not, see the [setup](setup.html) instructions for how to
change the working directory.

In the "data" directory, you should see 3 CSV files containing
station data:

```{r, eval=TRUE}divvy
Sys.glob("../data/*Stations*")
```

We will only use the Q4 data since it contains the most up-to-date
information about the Divvy bike stations.

## Loading the Divvy station data

The R environment is where all *variables and functions* are stored
and accessed.

Immediately after launching R, your R environment will be empty.

```{r, eval=FALSE}
ls()
```

We will create a new variable called "stations" by running
function "read.csv", and assigning the result of "read.csv" to
"stations":

```{r, eval=FALSE}
stations <- read.csv(file = "../data/Divvy_Stations_2016_Q4.csv",
                     stringsAsFactors = FALSE)
```

What does this function do, and what are the function arguments "file"
and "stringsAsFactors" for? We can easily find the answers to these
questions by accessing the R help documents for this function:

```{r, eval=FALSE}
help(read.table)
```

## A first peek at the Divvy station data 

What kind of variable is "stations"? We can check this:

```{r, eval=FALSE}
class(stations)
```

It is a "data frame", which is R's main data structure for tables.

Peek at the first few rows of the table:

```{r, eval=FALSE}
head(stations)
```

Peek at the last few rows of the table:

```{r, eval=FALSE}
tail(stations)
```

See the "README.txt" file in the data directory for more information
about the columns in the table.

Use `summary` to quickly generate an overview of the table:

```{r, eval=FALSE}
summary(stations)
```

This is a general function that works for many types of variables.
 
How many stations are there?

```{r, eval=FALSE}
nrow(stations)
ncol(stations)
```

We can inspect individual rows of the table, e.g.,

```{r, eval=FALSE}
stations[100,]
```

We can also take a closer look at individual columns of the table,
e.g., the "dpcapacity" column:

```{r, eval=FALSE}
class(stations[["dpcapacity"]])
class(stations$dpcapacity)
x <- stations$dpcapacity
summary(x)
table(x)
quantile(x,seq(0,1,0.1))
```

It is interesting that a few of the Divvy bike stations are much
larger than all the others. Where are these stations? 

```{r, eval=FALSE}
subset(stations,dpcapacity >= 40)
```

Alternatively, we can sort the table rows and inspect the top rows:

```{r, eval=FALSE}
rows     <- order(stations$dpcapacity,decreasing = TRUE)
stations <- stations[rows,]
head(stations,n = 8)
```

## Quickly accessing stations in the table by their name

It is often useful to access rows of a table by their name instead of
using a number. This can be done by setting the row names in the data
frame.

For example, this is one of the Divvy stations on the University of
Chicago campus:

```{r, eval=FALSE}
uofcstn <- "University Ave & 57th St"
i       <- which(stations$name == uofcstn)
stations[i,]
```

Let's now add row names:

```{r, eval=FALSE}
rownames(stations) <- stations$name
stations           <- stations[-2]
head(stations)
```

Note that I removed the second column in the table because it is no
longer needed.

Once we have row names, we can more easily access the row
corresponding to the Divvy bike station on the university campus:

```{r, eval=FALSE}
stations[uofcstn,]
```

## Save your session

This is a good time to save your current R environment just in case
you accidentally quit or lose your connection to the cluster.

```{r, eval=FALSE}
save.image(file = "workshop.RData")
```

## Recap

In this first episode, we explored some basic commands for loading
data from a text file into R, inspecting data in a data frame, and
making a few useful changes to the data.

In [the next episode](trips.html), we will explore these same
techniques, but on a larger scale, using the Divvy trip data.
