---
title: "Importing and inspecting the Divvy trip data"
author: "Peter Carbonetto"
output: html_document
---

<!-- Define defaults shared by all workflowr files. -->
```{r read-chunk, include=FALSE, cache=FALSE}
knitr::read_chunk("chunks.R")
```

<!-- Update knitr chunk options -->
```{r knitr-opts-chunk, include=FALSE}
```

<!-- Insert the date the file was last updated -->
```{r last-updated, echo=FALSE, results="asis"}
```

<!-- Insert the code version (Git commit SHA1) if Git repository
 exists and R package git2r is installed -->
```{r code-version, echo=FALSE, results="asis"}
```

For this part of the lesson, I'm assuming you have worked through the
code in the [previous episode](stations.html), and your R session is
still open. If you have saved your R session using `save.image`, you
can load it by running `load("workshop.RData")`.

<br>

## Installing the data.table package

Before, we used the `read.csv` function to read in the data. However,
if we try to use it for the Divvy Trip data, we will see that it is
very slow:

```{r, eval=FALSE}
trips <- read.csv(file = "../data/Divvy_Trips_2016.csv",
                  stringsAsFactors = FALSE,nrows = 1e5)
```

On your computer, it probably took a few seconds to read in the first
100,000 lines, but there is a total of 3.6 million lines! Instead, we
will use a function implemented in the `data.table` package that is
much faster.

Retrieve and install the package from CRAN:

```{r, eval=FALSE}
install.packages("data.table")
```

Load the package functions into your environment:

```{r, eval=TRUE}
library(data.table)
```

What functions were loaded? What can you do with this package? Use
this command to access the package documentation:

```{r, eval=FALSE}
help(package = data.table)
```

Package "vignettes" are also a great way to learn about a package,
e.g.,

```{r, eval=FALSE}
vignette("datatable-intro")
```

## Where are the packages?

*Add text here.*

## Use data.table to load the Divvy trip data

```{r, eval=FALSE}
```

## Inspect the data

## Create some factor columns

*Add text here.*

## Convert dates and times.

*Add text and code here.*


