---
title: "Analyze U of C biking trends from the Divvy data"
author: "Peter Carbonetto"
output: html_document
---

<!-- Define defaults shared by all workflowr files. -->
```{r read-chunk, include=FALSE, cache=FALSE}
knitr::read_chunk("chunks.R")
```

<!-- Update knitr chunk options -->
```{r knitr-opts-chunk, include=FALSE}
```

<!-- Insert the date the file was last updated -->
```{r last-updated, echo=FALSE, results='asis'}
```

<!-- Insert the code version (Git commit SHA1) if Git repository
 exists and R package git2r is installed -->
```{r code-version, echo=FALSE, results='asis'}
```

## Visualizing the usage patterns of University of Chicago Divvy users

Earlier, when we were inspecting the University of Chicago trip data,
we noticed an interesting trend: the trips end up much more frequently
at some stations than others. 

```{r, eval=FALSE}
summary(trips.uofc$to_station_name)
```

What could explain this pattern? One hypothesis is that geography can
explain this trend---to investigate, let's visualize these data.

First, create a variable storing counts of the number of trips by
arrival station.

```{r, eval=FALSE}
uofc.arrival.counts <- as.vector(table(trips.uofc$to_station_name))
head(uofc.arrival.counts)
```

Note that most stations in the city have never been visited from
University of Chicago.

```{r, eval=FALSE}
summary(factor(uofc.arrival.counts == 0))
```

Next, let's create another variable with the (Euclidean) distance
between the U of C Divvy station and all other stations across the
city.

```{r, eval=FALSE}
uofcstn <- "University Ave & 57th St"
uofc.station.dists <-
  sqrt((stations[uofcstn,"latitude"] - stations$latitude)^2 +
       (stations[uofcstn,"longitude"] - stations$longitude)^2)
head(uofc.station.dists)
```

To compare the counts and distance, let's combine these two variables
into a single data frame. Once we have a data frame, we can plot the
relationship between these two variables using ggplot.

```{r, eval=FALSE}
uofc.arrivals <- data.frame(dist  = uofc.station.dists,
                            count = uofc.arrival.counts)
rownames(uofc.arrivals) <- rownames(stations)
head(uofc.arrivals)
```

Note that we are able to combine these data into one data frame only
because we were careful earlier to create factors with levels given in
the same order as the stations in the "stations" data frame.

Let's create a scatterplot comparing the two variables.

```{r, eval=FALSE}
ggplot(data    = uofc.arrivals,
       mapping = aes(x = dist,y = count)) +
  geom_point()
```

This plot isn't particularly informative because a few trip counts are
much larger than the others, and a few of the distances are much
smaller. This suggests that a log-scale will work better.

```{r, eval=FALSE}
ggplot(data    = uofc.arrivals,
       mapping = aes(x = dist,y = count)) +
  geom_point() +
  scale_x_log10() +
  scale_y_log10()
```

That's better, but we get a couple warnings because we are taking the
log of zero. Let's fix this by removing all points at zero.

```{r, eval=FALSE}
ggplot(data    = subset(uofc.arrivals,count > 0 & dist > 0),
       mapping = aes(x = dist,y = count)) +
  geom_point(shape = 4,color = "darkblue") +
  scale_x_log10() +
  scale_y_log10(breaks = c(1,10,100,1000))
```

That's a big improvement. One final adjustment: let's add more
informative axis labels. And then we can save the plot for later.

```{r, eval=FALSE}
plot3 <- ggplot(data    = subset(uofc.arrivals,count > 0 & dist > 0),
                mapping = aes(x = dist,y = count)) +
  geom_point(shape = 4,color = "darkblue") +
  scale_x_log10() +
  scale_y_log10(breaks = c(1,10,100,1000)) +
  labs(x = "distance from University & 57th",
       y = "number of arrivals")
print(plot3)
```

This plot clearly shows that the number of trips to each station drops
off linearly with distance from the station (on the log-scale). It
would be interesting to see whether a similar trend occurs city-wide.

### Analyzing the linear relationship between two variables

Let's now analyze the relationship between trip counts and geographic
distance a little more rigorously. In particular, let's analyze the
linear relationship between these two quantities. The swiss army knife
for linear regression in R is the "lm" function. The "formula"
argument specifies the regression model. Here, we model log-counts as
a linear function of log-distances:

```{r, eval=FALSE}
fit <- lm(data    = subset(uofc.arrivals,count > 0 & dist > 0),
          formula = log10(count) ~ log10(dist))
```

As before, we can get a detailed summary of the linear regression
using the "summary" function:

```{r, eval=FALSE}
class(fit)
summary(fit)
```

Among other things, the summary tells us that about 50% of the
variation in trip arrival counts is explained by geographic distance.

We can also add the fitted regression line to the scatterplot we
created earlier:

```{r, eval=FALSE}
b <- coef(fit)
plot3 + geom_abline(intercept = b["(Intercept)"],slope = b["log10(dist)"],
                    color = "red",linetype = "dotted")
```

## Recap

This workshop was only a very brief tour of data analysis in R. To
learn more, please see the "Great resources for R" section at the
bottom of the [home page](index.html).
